---
title: "HIV Microbiome Analyses - Patel et al"
author: "Sweta Patel"
date: "5/15/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**General order of steps**
1. Basic demographics
2. Alpha diversity analyses
3. Transform data (CLR) and do Beta-diversity analyses (PCoA plots, PERMANOVA)
  #UPDATE 5/5: the order covariables are listed in PERMANOVA affects the results
4. Generate relative abundance plots using NON-transformed data
5. UPDATE 5/21: Filter on mean relative abundance of 50 and remove NOS species
6. Differential abundance analyses using Maaslin and FILTERED data
    #set Q value to 0.1-0.2
7. Logistic regression analyses (Aim 2)

*Note: these analyses utilize a pruned and decontaminated phyloseq object containing a cleaned taxtable and new variables created from the original variables recorded in REDCap. All coding can be found in the "Fog KAIJU 10-6-23" file.*

```{r include=FALSE}
#Loading packages 
library(plyr)
library(dplyr)
library(ggplot2)
library(phyloseq)
library(tidyr)
library(vegan)
library(metagenomeSeq)
library(httr)
library(gridExtra)
library(data.table)
library(RColorBrewer)
library(scales)
library(gsubfn)
library(readxl)
library(microbiome)
library(MASS)
library(Maaslin2)
library(cowplot)
```

```{r include=FALSE}
set.seed(1234)
#can't set wd in RMD file; entered below code into console directly
# setwd("OneDrive - Duke University/Fogarty coding/Sequencing/")
```

### Importing and cleaning data

```{r, results = FALSE}
pruned <- readRDS("/Users/swetapatel/OneDrive - Duke University/Fogarty coding/Sequencing/phy.pruned05092022.rds")
#confirm the taxtable, etc looks good:
test <- as.data.frame(as(tax_table(pruned),"matrix"),stringsAsFactors=FALSE)  #success!
meta_prune <- data.frame(sample_data(pruned))
meta_prune <- within(meta_prune, rm(sib_weight, sib_sex, sib_height, sib_muac, sib_race, sib_bfeed_1,
                                  sib_bfeed_2, sib_uri_recent, sib_uri_current, sib_pcr_1, sib_pcr_2,
                                  sib_bcg, sib_hepb, sib_clinic, sib_clinic_dx___5, sib_clinic_dx_oth,
                                  sib_hosp, sib_meds, sib_meds_name, sibmo, sibyr, sib_vl1_weeks, sib_vl_weeks2,
                                  sib_med_days, sib_dpt, sib_pcv, sib_rota, sib_polio, sib_measles, sib_age, samp_storage,
                                  enr_clinic, enr_clinic_other, mat_np))
remove(test)
```

### Demographics of participants with sequencing results

N= 272 participants with sequencing results (of 309 enrolled participants)

```{r}
child <- subset(meta_prune, subject == "child") #n=143
table(child$hiv)  #50 HUU kids, 49 HEU kids, 44 HIV+ kids

#Age
hist(child$age)
summary(child$age)
tapply(child$age, child$hiv, summary)
kruskal.test(child$age, child$hiv)

#Sex
table(child$sex)  #74 girls
table(child$sex, child$hiv)
prop.table(table(child$sex, child$hiv), 2)
summary(table(child$sex, child$hiv))

#Maternal age
hist(child$mat_age2)  #looks quite normal, but will still present median and IQR for consistency
summary(child$mat_age2)
tapply(child$mat_age2, child$hiv, summary)
kruskal.test(child$mat_age2, child$hiv)

#Season
table(child$season) 
table(child$season, child$hiv)
prop.table(table(child$season, child$hiv), 2)
summary(table(child$season, child$hiv))

#Maternal education
table(child$mat_educ)
table(child$mat_educ2)
table(child$mat_educ2, child$hiv)
prop.table(table(child$mat_educ2, child$hiv), 2)
summary(table(child$mat_educ2, child$hiv))
fisher.test(child$mat_educ2, child$hiv)

#Electricity
table(child$elec) #106 children have electricity
table(child$elec, child$hiv)
prop.table(table(child$elec, child$hiv), 2)
summary(table(child$elec, child$hiv))

#Wood
table(child$wood) #90 children in households using wood
table(child$wood, child$hiv)
prop.table(table(child$wood, child$hiv), 2)
summary(table(child$wood, child$hiv))

#Num household members
hist(child$hhsize)
summary(child$hhsize)
tapply(child$hhsize, child$hiv, summary)
kruskal.test(child$hhsize, child$hiv)

#Num children <5
hist(child$num_child)
summary(child$num_child)
tapply(child$num_child, child$hiv, summary)
kruskal.test(child$num_child, child$hiv)

#antibiotics
table(child$abx) #24 children received abx
table(child$abx, child$hiv)
prop.table(table(child$abx, child$hiv), 2)
summary(table(child$abx, child$hiv))

#Clinic in past 3 months
table(child$clinic)
table(child$clinic, child$hiv)
prop.table(table(child$clinic, child$hiv), 2)
summary(table(child$clinic, child$hiv))

#URI in past 1 month
table(child$uri_recent)
table(child$uri_recent, child$hiv)
prop.table(table(child$uri_recent, child$hiv), 2)
summary(table(child$uri_recent, child$hiv))

#URI currently (don't need both this and uri recent in table 1; need to choose)
table(child$uri_current)
table(child$uri_current, child$hiv)
prop.table(table(child$uri_current, child$hiv), 2)
summary(table(child$uri_current, child$hiv))

#Hospital in past 3 months -->> 4 children hospitalized, all 4 HIV+
  #*MENTION IN MANUSCRIPT
table(child$hosp)
table(child$hosp, child$hiv)
table(child$hosp_dx)  #1 with lower respiratory infx, 1 with gastroenteritis, 2 with other
table(child$hosp_dx_oth)  #malnutrition, sepsis + chronic suppurative OM

#PCV doses
table(child$pcv)
table(child$pcv, child$hiv)
prop.table(table(child$pcv, child$hiv), 2)
summary(table(child$pcv, child$hiv))
fisher.test(child$pcv, child$hiv)

#HiB
table(child$dpt)
table(child$dpt, child$hiv)
prop.table(table(child$dpt, child$hiv), 2)
summary(table(child$dpt, child$hiv))
fisher.test(child$dpt, child$hiv)

#breakdown of subjects (not for table 1, but for sequencing figure)
table(meta_prune$subject)
```

### HIV-specific variables
```{r}
chi_hiv <- subset(child, hiv == "Infected")
#restrict to just HIV variables for easier cleaning and analysis
chi_hiv <- chi_hiv[c("child_pcr_1", "child_vl_1", "child_pcr_2", "child_vl_2", "chi_cd4_num_1", 
                     "chi_cd4_perc_1", "chi_cd4_num_2", "chi_cd4_perc_2", "chi_cd4_weeks1", 
                     "chi_cd4_weeks2", "child_vl1_weeks", "child_vl2_weeks",
                     "child_arv", "child_arv_meds", "child_tmpsmx", "abx")]

#How many kids receiving TMP-SMX? NOTE: we did not include these kids in the abx category. Need to re-do analyses with this included?
table(chi_hiv$child_tmpsmx) #15 kids on TMP-SMX
#manual inspection: 10 kids on TMP-SMX without other abx exposure

#I want the most recent viral load values for my analysis. Will try creating a new variable to ID which value is more recent first:
#missing 3 vl1 and 14vl2's
chi_hiv$newest_vL <- NA
chi_hiv$newest_vL [chi_hiv$child_vl1_weeks < chi_hiv$child_vl2_weeks] <- 1
chi_hiv$newest_vL [chi_hiv$child_vl2_weeks < chi_hiv$child_vl1_weeks] <- 2
chi_hiv$newest_vL [chi_hiv$child_vl2_weeks < 0] <- 1  #we want to include the viral loads collected BEFORE enrollment where possible
chi_hiv$newest_vL [is.na(chi_hiv$child_vl_2)] <- 1    #if we are missing vl_2, we will use vl_1
chi_hiv$newest_vL [is.na(chi_hiv$child_vl_1) & is.na(chi_hiv$child_vl_2)] <- NA
table(chi_hiv$newest_vL)

#Next, can we pull the viral load that was most recent using our newest_vL variable?
  #source: https://stackoverflow.com/questions/67944727/creating-new-variable-by-selecting-column-based-on-value-of-another-column
  #Need to do this with rows that have values (aka need to exclude the NA viral load children for this to work)
#For future phyloseq analyses: will need to remove B-04, B-07, B-52
chi_viral <- subset(chi_hiv, !is.na(newest_vL))
chi_viral$Var1 <- chi_viral$child_vl_1
chi_viral$Var2 <- chi_viral$child_vl_2

chi_viral <- chi_viral %>% rowwise() %>%
  mutate(newest_vlnum = get(paste0('Var', newest_vL)))
#now we have a dataframe of the 41 children with viral load measurements and 1 column containing the most recent viral loads for each
summary(chi_viral$newest_vlnum)
table(chi_viral$newest_vlnum) #27 with viral suppression, 14 without viral suppression

#make new variable for viral suppression or not, with vL < 400 = suppressed
chi_viral$vl_suppr <- NA
chi_viral$vl_suppr [chi_viral$newest_vlnum < 400] <- "1"
chi_viral$vl_suppr [chi_viral$newest_vlnum >= 400] <- "0"
table(chi_viral$vl_suppr)

#REPEAT the above steps to get date associated with most recent viral load, most recent CD4, and date a/w most recent CD4
#Viral load collection date
chi_viral$Var1 <- chi_viral$child_vl1_weeks
chi_viral$Var2 <- chi_viral$child_vl2_weeks

chi_viral <- chi_viral %>% rowwise() %>%
  mutate(newest_vldate = get(paste0('Var', newest_vL)))   #now have info for when vL were checked in relation to enrollment

summary(chi_viral$newest_vldate)

chi_viral <- within(chi_viral, rm(Var1, Var2))

#CD4
  #missing CD4 data for B-45 and B-52
  #will focus on percentages since this is what is presented for kids
chi_hiv$newest_cd4 <- NA
chi_hiv$newest_cd4 [chi_hiv$chi_cd4_weeks1 < chi_hiv$chi_cd4_weeks2] <- 1
chi_hiv$newest_cd4 [chi_hiv$chi_cd4_weeks2 < chi_hiv$chi_cd4_weeks1] <- 2
chi_hiv$newest_cd4 [chi_hiv$chi_cd4_weeks2 < 0] <- 1  #we want to include the viral loads collected BEFORE enrollment where possible
chi_hiv$newest_cd4 [is.na(chi_hiv$chi_cd4_perc_2)] <- 1    #if we are missing cd4_2, we will use cd4_1
chi_hiv$newest_cd4 [is.na(chi_hiv$chi_cd4_perc_1)] <- 2    #one child is missing cd4_1 but has cd4_2
chi_hiv$newest_cd4 [is.na(chi_hiv$chi_cd4_perc_1) & is.na(chi_hiv$chi_cd4_perc_2)] <- NA
table(chi_hiv$newest_cd4)

#create new dataframe without B-45 and B-52 (aka new df with children missing CD4 data removed)
chi_cd4 <- subset(chi_hiv, !is.na(newest_cd4))
chi_cd4$Var1 <- chi_cd4$chi_cd4_perc_1
chi_cd4$Var2 <- chi_cd4$chi_cd4_perc_2

chi_cd4 <- chi_cd4 %>% rowwise() %>%
  mutate(newest_cd4perc = get(paste0('Var', newest_cd4)))

summary(chi_cd4$newest_cd4perc)
table(chi_cd4$newest_cd4perc)

#make new variable for immune competent CD4 or not, with CD4 percentage *** = immune competent
  #See sub-analyses file for code

#arvs
table(chi_hiv$child_arv)
table(chi_hiv$child_arv_meds)
```

### Number of reads by subject class and HIV status

*using pruned and agglomerated reads*

```{r}
summary(meta_prune$paired.reads)   #Median (IQR) of 73,166 (29,605; 217,115) reads per sample
#mean 183,726 reads per sample
hist(meta_prune$paired.reads)

#Did maternal samples contain more reads than child or sib reads? --> the opposite! Why?
#children = petri dishes
tapply(meta_prune$paired.reads, meta_prune$subject, summary)
kruskal.test(meta_prune$paired.reads, meta_prune$subject)

#number of reads
hist(child$paired.reads)
summary(child$paired.reads)
tapply(child$paired.reads, child$hiv, summary)
kruskal.test(paired.reads ~ hiv, data = child) 

#Sum of sequences
sum(sample_sums(pruned))     #35,101,748 sequences obtained 
mean(sample_sums(pruned))    #Mean of 129,050.5 sequences per sample
median(sample_sums(pruned))  #Median of 45,235 sequences per sample
summary(sample_sums(pruned)) #IQR of sequencing reads per sample: 15730, 146428
nsamples(pruned)        #272 NP samples 
ntaxa(pruned) #6709 taxa
```

### Alpha diversity measures

*Create dataframe*

```{r}
#Create dataframe with diversity indices for each specimen using PRUNED pseq object
diversity <- estimate_richness(pruned, measures = c("Shannon", "Simpson", "Chao1", "Observed"))

setDT(diversity, keep.rownames = TRUE)[]
colnames(diversity) <- c("sample_id", "Observed", "Chao1", "se.chao1", "Shannon", "Simpson")
rownames(diversity) <- diversity$sample_id
diversity <- merge(diversity, meta_prune, by="sample_id")
```

#### Observed species

```{r}
#How many species on average per sample? Did that differ by subject or by HIV status (among children only)?
hist(diversity$Observed)
summary(diversity$Observed) 
sd(diversity$Observed)
#Median (IQR) # spec: 106 (53, 260)
#Mean (SD): 216 (300)

#what does it look like by subject? Moms have significantly fewer species per sample
tapply(diversity$Observed, diversity$subject, summary)   
kruskal.test(Observed ~ subject, data = diversity)

#what about by HIV status of the child? No signif diff
div_child <- subset(diversity, subject == "child")
tapply(div_child$Observed, div_child$hiv, summary)
kruskal.test(Observed ~ hiv, data = div_child) 
```

#### Shannon

```{r}
#Is our SDI data normally distributed? 
shapiro.test(diversity$Shannon)  #technically no, because p<0.05; looks pretty good on histogram but QQ is wonky
histogram(diversity$Shannon)
qqnorm(diversity$Shannon, pch = 1, frame = FALSE)
qqline(diversity$Shannon, col = "steelblue", lwd = 2)

#Shannon diversity median and IQR for all samples
summary(diversity$Shannon)
sd(diversity$Shannon)
#Median (IQR) SDI: 1.85 (1.51, 2.26)
#Mean (SD) 1.93 (0.66)  

#what does it look like by subject? 
tapply(diversity$Shannon, diversity$subject, summary)
kruskal.test(Shannon ~ subject, data = diversity) #child signif lower (p=0.01)
#Since our data is normal-ish in distribution, what does the ANOVA look like?
subj.aov <- aov(Shannon ~ subject, data = diversity)
summary(subj.aov)   #P value 0.007; SIGNIFICANT
table(diversity$subject)

#what about by HIV status of the child?
tapply(div_child$Shannon, div_child$hiv, summary)
kruskal.test(Shannon ~ hiv, data = div_child)   #no significant difference by HIV status using KW test (P=0.71)
  #ANOVA
hiv.aov <- aov(Shannon ~ hiv, data = div_child)
summary(hiv.aov)  #P=0.37, not significant

#Age of the child
div_age <- lm(Shannon ~ age, data = div_child)
summary(div_age)  #p=0.98

#Wood smoke exposure
tapply(div_child$Shannon, div_child$wood, summary)
wilcox.test(Shannon ~ wood, data = div_child)   #p=0.30
t.test(Shannon ~ wood, data = div_child)        #P=0.73

#Season
tapply(div_child$Shannon, div_child$season, summary)
wilcox.test(Shannon ~ season, data = div_child)   #p=0.69
t.test(Shannon ~ season, data = div_child)  #P=0.70

#Receipt of abx in prior 3 months 
tapply(div_child$Shannon, div_child$abx, summary)
wilcox.test(Shannon ~ abx, data = div_child)   #p=0.86
t.test(Shannon ~ abx, data = div_child)   #P=0.71

#Current URI
tapply(div_child$Shannon, div_child$uri_cur2, summary)
wilcox.test(Shannon ~ uri_cur2, data = div_child) #p=0.65
t.test(Shannon ~ uri_cur2, data = div_child) #p=0.37

#Recent URI
tapply(div_child$Shannon, div_child$uri_rec2, summary)
wilcox.test(Shannon ~ uri_rec2, data = div_child) #p=0.30
t.test(Shannon ~ uri_rec2, data = div_child) #p=0.31

#Household members
summary(div_hh <- lm(Shannon ~ hhsize, data = div_child)) #p=0.47

#MULTIVARIABLE MODEL
#Need to set reference for HIV variable
class(div_child$hiv) #character; need to make factor to relevel
div_child$hiv <- as.factor(div_child$hiv)
div_child$hiv <- relevel(div_child$hiv, ref = "Unexposed")
summary(div_mv <- lm(Shannon ~ age + hiv + wood + season + abx + uri_rec2, data = div_child))
summary(div_mv <- lm(Shannon ~ age + hiv + wood + season + abx + uri_rec2 + hhsize, data = div_child))
```

#### Chao1

```{r}
#Is our Chao1 data normally distributed? 
shapiro.test(diversity$Chao1)  #no, because p<0.05; also looks skewed on histogram and QQ is wonky
histogram(diversity$Chao1)
qqnorm(diversity$Chao1, pch = 1, frame = FALSE)
qqline(diversity$Chao1, col = "steelblue", lwd = 2)

#What does it look like if we log transform it? Will this allow for parametric testing?
diversity$Chao1_log <- log(diversity$Chao1)
shapiro.test(diversity$Chao1_log) 
histogram(diversity$Chao1_log)
qqnorm(diversity$Chao1_log, pch = 1, frame = FALSE)
qqline(diversity$Chao1_log, col = "steelblue", lwd = 2)
  #Looks very normal! Can use t.test and linear regression on transformed data
div_child <- subset(diversity, subject == "child")

#Chao1 median and IQR for all samples
summary(diversity$Chao1)
sd(diversity$Chao1)
#Median (IQR) Chao1: 106 (53, 260)
#Mean (SD) 216 (300.4)

#what does it look like by subject? 
tapply(diversity$Chao1, diversity$subject, summary)
tapply(diversity$Chao1_log, diversity$subject, summary) #log mean of moms is lower
summary(aov(Chao1_log ~ subject, data = diversity)) #significant difference (p=0.03)

#what about by HIV status of the child?
tapply(div_child$Chao1, div_child$hiv, summary)
tapply(div_child$Chao1_log, div_child$hiv, summary)
summary(aov(Chao1_log ~ hiv, data = div_child)) #no significant difference by HIV status using log values and ANOVA (p=0.2)

#Age of the child
summary(div_age <- lm(Chao1_log ~ age, data = div_child)) #p=0.38

#Wood smoke exposure
tapply(div_child$Chao1, div_child$wood, summary)
t.test(Chao1_log ~ wood, data = div_child)    #p=0.93

#Season
tapply(div_child$Chao1, div_child$season, summary)
t.test(Chao1_log ~ season, data = div_child)    #p=0.43

#Receipt of abx in prior 3 months 
tapply(div_child$Chao1, div_child$abx, summary)
t.test(Chao1_log ~ abx, data = div_child)    #p=0.92

#Current URI  #p=0.06
tapply(div_child$Chao1_log, div_child$uri_cur2, summary)
t.test(Chao1_log ~ uri_cur2, data = div_child)

#Recent URI 
tapply(div_child$Chao1_log, div_child$uri_rec2, summary)
t.test(Chao1_log ~ uri_rec2, data = div_child)  #p=0.69

#Household members
summary(div_hh <- lm(Chao1_log ~ hhsize, data = div_child)) #p=0.42

#MULTIVARIABLE MODEL
#log with HUU as reference value
class(div_child$hiv) #character; need to make factor to relevel
div_child$hiv <- as.factor(div_child$hiv)
div_child$hiv <- relevel(div_child$hiv, ref = "Unexposed")
summary(div_mv <- lm(Chao1_log ~ age + hiv + wood + season + abx + uri_rec2, data = div_child))
summary(div_mv <- lm(Chao1_log ~ age + hiv + wood + season + abx + uri_cur2, data = div_child))
summary(div_mv <- lm(Chao1_log ~ age + hiv + wood + season + abx + uri_rec2 + hhsize, data = div_child))
```

#### Alpha diversity boxplots

- Helpful resource for boxplots: http://www.sthda.com/english/wiki/ggplot2-box-plot-quick-start-guide-r-software-and-data-visualization

```{r}
#make new variable with HIV status abbreviated
div_child$hiv2 <- NA
div_child$hiv2 [div_child$hiv == "Infected"] <- "CLWH"
div_child$hiv2 [div_child$hiv == "Exposed Uninfected"] <- "HEU children"
div_child$hiv2 [div_child$hiv == "Unexposed"] <- "HUU children"
table(div_child$hiv2)
#matches up with table(relative_df$hiv)
div_child$hiv2 <- as.factor(div_child$hiv2)
div_child$hiv2 <- reorder(div_child$hiv2, new.order=c("CLWH", "HEU children", "HUU children"))

#Shannon boxplot 
s1a2 <- ggplot(div_child, aes(x=hiv2, y=Shannon, fill=hiv2)) + 
  geom_boxplot() + labs(x="HIV status", y = "Shannon index")+
  scale_fill_manual(values=c('#800000FF','#155F83FF', '#FFA319FF')) +
  theme_classic() +
  theme(
    legend.text = element_text(size = 10),
    legend.position = "none",
    legend.title = element_blank(),
    axis.text = element_text(size = 10),
    axis.title.y = element_text(size = 12),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank())

#Chao1 by HIV (to create legend we will use for our combined figure)
s1b <- ggplot(div_child, aes(x=hiv2, y=Chao1, fill=hiv2)) + 
  geom_boxplot() + labs(x="HIV status", y = "Chao1 richness")+
  scale_fill_manual(values=c('#800000FF','#155F83FF', '#FFA319FF')) +
  theme_classic() +
  theme(
    legend.text = element_text(size = 10),
    legend.position = "right",
    legend.title = element_blank(),
    axis.text = element_text(size = 10),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.text.x = element_text(size=10, angle = 45, vjust = 1, hjust = 1, colour = "black"))

#Chao1 boxplot with no legend
s1b2 <- ggplot(div_child, aes(x=hiv2, y=Chao1, fill=hiv2)) + 
  geom_boxplot() + labs(x="HIV status", y = "Chao1 richness")+
  scale_fill_manual(values=c('#800000FF','#155F83FF', '#FFA319FF')) +
  theme_classic() +
  theme(
    legend.text = element_text(size = 10),
    legend.position = "none",
    legend.title = element_blank(),
    axis.text = element_text(size = 10),
    axis.title.y = element_text(size = 12),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank())

#Get legend only to make it easier to size all parts of the figure
legend <- get_legend(s1b)

#combine the versions using cowplot to create supplemental Figure S1
prow2 <- plot_grid(s1a2, s1b2, labels = "AUTO", nrow = 1, ncol = 2, align = "h", axis = "b", rel_widths = c(1,1.1))
FigS1_v2 <- plot_grid(prow2, legend, rel_widths = c(2, 0.5))

png(file="/Users/swetapatel/OneDrive - Duke University/Fogarty coding/FigS1_v2.png",
    width = 7, height = 3.5, units = 'in', res = 600)
FigS1_v2
dev.off()

#save as .eps file to comply with Lancet Microbe requirements
ggsave(file="/Users/swetapatel/OneDrive - Duke University/Fogarty coding/FigS1_v2.eps", width = 7, height = 3.5, units = "in")
FigS1_v2
dev.off()

#clean up environment
remove(div_age, div_mv, hiv.aov, m1, subj.aov)
```
- source for saving as .eps: https://stackoverflow.com/questions/5142842/export-a-graph-to-eps-file-with-r

### Beta diversity measures

*In order to compare overall composition by HIV status, we will first CLR transform our data and use euclidean distances for comparison*

- Sources: https://rdrr.io/github/microbiome/microbiome/man/transform.html
  https://microbiome.github.io/tutorials/Preprocessing.html

```{r}
pruned_clr <- microbiome::transform(pruned, "clr")
chi_clr = subset_samples(pruned_clr, subject == "child")
```

#### Beta diversity plots

- will save in .eps format for consistency with Lancet Microbe requirements

```{r}
#First: create new HIV variable as factor, then relevel
table(child$hiv)
child$hiv2 <- NA
child$hiv2 [child$hiv == "Infected"] <- "CLWH"
child$hiv2 [child$hiv == "Exposed Uninfected"] <- "HEU children"
child$hiv2 [child$hiv == "Unexposed"] <- "HUU children"
table(child$hiv2)
class(child$hiv2)
child$hiv2 <- as.factor(child$hiv2)
child$hiv2 <- reorder(child$hiv2, new.order=c("CLWH", "HEU children", "HUU children"))

#Import back into pseq (source: https://jacobrprice.github.io/2017/08/26/phyloseq-to-vegan-and-back.html)
sample_data(chi_clr) <- as.data.frame(child)
#now see if it worked:
bloop <- data.frame(sample_data(chi_clr))
table(bloop$hiv2)
#success!
remove(bloop)

#Now ordinate and create PCOA plot
clr_pcoa <- ordinate(chi_clr, method = "PCoA", distance = "euclidean")
p1 <- plot_ordination(chi_clr, clr_pcoa, color = "hiv2") +
  geom_point(size = 0.005) + theme_classic() + theme(
    plot.title = element_text(size = 14, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    legend.position = "right",
    legend.background = element_blank(),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)) +
  scale_color_manual(values=c('#800000FF','#155F83FF', '#FFA319FF'))+
  stat_ellipse(aes(linetype=hiv2), geom="polygon", alpha=0, type="t", level=0.8, size=0.5) +
  xlab("PC1 (14.4%)") + ylab("PC2 (8.9%)") +
  scale_linetype_manual(values= c("dashed", "dashed", "dashed"))
```

#### Permanova

- NOTE: since initial analysis, adonis has been depreciated so all analyses here updated with adonis2

```{r}
#Create OTU table
otu <- as.data.frame(otu_table(chi_clr, taxa_are_rows=TRUE))

#Need species in columns and observations in rows: transpose using t() function and save as a dataframe
  #Source: https://github.com/edamame-course/2015-tutorials/blob/master/final/Ashley_Intro_to_R.md
otu_trans <- data.frame(t(otu))
child <- data.frame(sample_data(chi_clr))

#Calculate distance and save as a matrix
BC.dist=vegdist(otu_trans, method="euclidean")

#HIV
# adonis(BC.dist ~ hiv, data = child, permutations = 1000)
  #Significant difference by HIV status among children (P = 0.043, R^2 = 0.019)
adonis2(BC.dist ~ hiv, data = child, permutations = 1000)
  #Significant difference by HIV status remains: R2 = 0.019, P = 0.039

#HIV + age (per 6/9/23 MK discussion to account for age in all PERMANOVAs)
#NOTE re: adonis2 'by' function: "by = "terms" will assess significance for each term (sequentially from first to last), setting by = "margin" will assess the marginal effects of the terms (each marginal term analysed in a model with all other variables)"
#Source: https://search.r-project.org/CRAN/refmans/vegan/html/adonis.html
#Previously, we used the by = "terms" function and ordered terms in multivariable models based on the R2 from each univariable model. However, it is worth exploring the by = "margin" so we can look at the marginal effects of all variables together.
# adonis(BC.dist ~ hiv + age, data = child, permutations = 1000)
  #Significant difference by HIV status among children (P = 0.042, R^2 = 0.019)
    #Significant difference by age as well (P = 0.03, R^2 = 0.012)

#NEW adonis2 code with by = "terms" and by = "margins"
adonis2(BC.dist ~ hiv + age, data = child, permutations = 1000, by = "terms")
  #Significant difference by HIV status among children (P = 0.036, R2 = 0.019) and by age (P = 0.023, R2 = 0.012
adonis2(BC.dist ~ hiv + age, data = child, permutations = 1000, by = "margin")
  #results are pretty consistent: HIV: p = 0.043, R2 = 0.019; age: p = 0.02, R2 = 0.12

#Age
adonis2(BC.dist ~ age, data = child, permutations = 1000)
  #Significant difference in overall composition by age (P = 0.03, R2 = 0.012)

#Season
  #Significant difference in overall composition by season (P < 0.001, R2 = 0.016)
adonis2(BC.dist ~ season, data = child, permutations = 1000)
  #Season + age (per 6/9 MK discussion to account for age in all PERMANOVAs)
adonis2(BC.dist ~ season + age, data = child, permutations = 1000, by = "terms")
adonis2(BC.dist ~ season + age, data = child, permutations = 1000, by = "margin")
#Significant difference by season (P = 0.002, R^2 = 0.016) and age (P = 0.03, R2 = 0.011) by "terms"
#Consistent by "margin": season (P = 0.004, R^2 = 0.016) and age (P = 0.03, R2 = 0.011)

#Wood
adonis2(BC.dist ~ wood, data = child, permutations = 1000)
  #No significant difference by wood smoke exposure among children (P = 0.12)
#Wood + age (per 6/9 MK discussion to account for age in all PERMANOVAs)
adonis2(BC.dist ~ wood + age, data = child, permutations = 1000, by = "terms")
#No signif difference by wood smoke exposure (P = 0.09) but age remains signif: (P = 0.02, R^2 = 0.012)
adonis2(BC.dist ~ wood + age, data = child, permutations = 1000, by = "margin") #consistent

#Abx
adonis2(BC.dist ~ abx, data = child, permutations = 1000)
  #Trend but no significant difference in overall composition by abx exposure (P = 0.08)
#Abx + age (per 6/9 MK discussion to account for age in all PERMANOVAs)
adonis2(BC.dist ~ abx + age, data = child, permutations = 1000, by = "terms")
#Trend but no significant difference in overall composition by abx exposure (P = 0.06) and age remains significant; same for by = "margin"
adonis2(BC.dist ~ abx + age, data = child, permutations = 1000, by = "margin")

#Recent URI
adonis2(BC.dist ~ uri_rec2, data = child, permutations = 1000)
#Significant difference in composition among children with recent URI (P = 0.008, R2=0.015)
  #Recent URI + age (per 6/9 MK discussion to account for age in all PERMANOVAs)
adonis2(BC.dist ~ uri_rec2 + age, data = child, permutations = 1000, by = "terms")
adonis2(BC.dist ~ uri_rec2 + age, data = child, permutations = 1000, by = "margin")
#Significant difference in composition by recent URI sx status (R2 = 0.015, p=0.004) and age (R2 = 0.11, p = 0.04) using both "terms" and "margin"

#Multivariable
  #NOTE: order of variables affects results when using by = "terms" per link above
#start with model that lists covariables by effect size from univariable analysis (signif variables only)
  #HIV R2 = 0.019, season R2 = 0.016, recent URI R2=0.015, age: R2 = 0.012
adonis2(BC.dist ~ hiv + season + uri_rec2 + age, data = child, permutations = 1000, by = "terms")
pm <- adonis2(BC.dist ~ hiv + season + uri_rec2 + age, data = child, permutations = 1000, by = "margin")
  #In a multivariable model, HIV status (P=0.04, R2=0.019), season (P=0.002, R2=0.016), recent URI (P=0.01, R2=0.013), and age (P=0.04, R2=0.011) were a/w differences in overall microbiome composition using both by = options (above R2 and p values are from by = "margin")
```

Statistical reviewer from Lancet Microbe requested 95% CI for each R2 we presented. From web review appears the most straightforward way to obtain these values is through bootstrapping:
https://stats.stackexchange.com/questions/175026/formula-for-95-confidence-interval-for-r2
https://rstudio-pubs-static.s3.amazonaws.com/288625_b0cf85c5be974eb795cef778535289d8.html

General tutorial on bootstrapping:
https://stats.oarc.ucla.edu/r/faq/how-can-i-generate-bootstrap-statistics-in-r/

```{r}
library(boot)
test <- function(d,i) {
  test2 <- d[i,]
  return(adonis2(BC.dist ~ hiv + season + uri_rec2 + age, data = child, permutations = 1000, by = "margin"))
}

set.seed(1234)
boota <- boot(child, test, R=1000)
boota

```